Certified Symbolic Finite Transducers: Formalization and Applications to String Manipulation

### 2025-07-20 Paper/main.tex
minor
update the paper

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

\title{Certified Symbolic Transducer with Applications in String Solving} %TODO Please add

@@

\author{Shuanglong Kan}{Barkhausen Institut, Germany \and \url{https://github.com/ShlKan} }{shuanglongkan@gmail.com}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\authorrunning{S. Kan} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{16th Conference on
Interactive Theorem Proving (ITP 2025)}
\EventShortTitle{ITP 2025}
\EventAcronym{ITP}
\EventYear{2025}
\EventDate{July 24--27, 2025}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Finite Automata (FAs) are fundamental components in the domains of programming languages. For instance, regular expressions, which are pivotal in languages such as JavaScript and Python, are frequently implemented using FAs.
%
Finite Transducers (FTs) extend the capabilities of FAs by enabling the transformation of input strings into output strings, thereby providing a more expressive framework for operations that encompass both recognition and transformation. Despite the various formalizations of FAs in proof assistants such as Coq and Isabelle/HOL, these formalizations often fall short in terms of applicability to real-world scenarios. A significant limitation of classical FAs and FTs is that transition labels are typically defined as a single character from a finite alphabet. However, in practical applications, the alphabet of an FA or FT can be enormously large or even \emph{infinite}. The classical approach to formalizing transitions can result in transition explosion, leading to critical performance bottlenecks of FA and FT operations.

A more pragmatic approach involves the formalization of symbolic FAs \cite{cav/DAntoniV17} and FTs \cite{VeanesHLMB12Transducer}, where transition labels are symbolic and potentially infinite. While the formalization of symbolic FAs has been explored in the work of CertiStr \cite{cpp/KanLRS22}, the formalization of symbolic FTs in interactive proof assistants remains largely unexplored due to the increased complexity challenges.
%
In this paper, we aim to formalize symbolic FTs within the Isabelle/HOL framework. This formalization is refinement-based and is designed to be extensible with various symbolic representations of transition labels. To assess its performance, we applied the formalized symbolic FTs to an SMT string solver for modeling replacement operations. The experimental results indicate that the formalized symbolic transducer can efficiently and effectively solve string constraints with replacement operations.

\end{abstract}

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite transducers in Isabelle/HOL proof assistant.
\item We leverage the refinement framework to create an extensible SFT formalization that is capable of accommodates various boolean algebras, ensuring adaptability to future transition label developments.
\item We develop an interval algebra with efficient set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization through its application to string constraint solving, specifically in modeling replacement operations with verified correctness guarantees.
\end{enumerate}

@@

For a word $w = a_0,\ldots, a_{n-1}$, it is accepted by $\sigma$ if and only if $a_i \in \llbracket\phi_i \rrbracket$ for $0 \leq i \leq n - 1$. When $w$ is accepted, run $\sigma$ generates an output sequence $w'= f_0~a_0, \ldots, f_{n-1}~a_{n-1}$. We define:
\begin{itemize}
\item  $(a_0,(f_0~a_0)), \ldots, (a_{n-1},(f_{n-1}~a_{n-1}))$ a \emph{trace},
\item $a_0,\ldots,a_{n-1}$ the \emph{input} of the trace and $(f_0~a_0), \ldots, (f_{n-1}~a_{n-1})$ the \emph{output} of the trace.
\end{itemize}

If $a_0,\ldots,a_{n-1}$ is accepted by an accepting run in $\mathcal{T}$, we say that the trace is an accepting trace of $\mathcal{T}$.
For a trace $\pi$, we denote its input as $\mathit{in}(\pi)$ and its output as $\mathit{out}(\pi)$.
%
Given an SFT $\mathcal{T}$ and a word $w$, we define the \emph{product} operation of $\mathcal{T}$ and $w$ (denoted as $\mathcal{T}\times\{w\}$) as the set of outputs generated by $\mathcal{T}$ with input $w$. More precisely, 
\[
\mathcal{T}\times\{w\} = \{w'\mid \exists \pi.~\pi \text{ is an accepting trace of } \mathcal{T} \land in(\pi) = w \land out(\pi) = w'\}.
\]

To make the operation \emph{product} more general, we extend the operation to an SFT and a set of input words represented by a regular language, which can be denoted by an SFA $\mathcal{A}$. More precisely,
\[
\mathcal{T}\times \mathcal{A} = \{w'\mid \exists w.~w\in \mathcal{L}(\mathcal{A})\land w' \in \mathcal{T}\times\{w\}\}, \text{ where }\mathcal{L}(\mathcal{A})\text{ denotes the language of }\mathcal{A}.
\]

\begin{figure}[hbt!]
  \centering
  \begin{tikzpicture}[shorten >=1pt, node distance=3.5cm, on grid, auto]
    \node[state, initial] (q_0)   {$q_0$}; 
    \node[state] (q_1) [right=of q_0] {$q_1$}; 
    \node[state, accepting] (q_2) [right=of q_1] {$q_2$};

    \path[->] 
    (q_0) edge node {[$'\texttt{a}'-'\texttt{z}'$],$f_1$} (q_1)
    (q_1) edge node {[$'\texttt{A}'-'\texttt{Z}'$],$f_2$} (q_2)
    (q_1) edge[loop above] node {[$'\texttt{a}'-'\texttt{z}'$],$f_1$} (q_1)
    (q_2) edge[loop above] node {[$'\texttt{A}'-'\texttt{Z}'$],$f_2$} (q_2);
 \end{tikzpicture}
    \caption{An example of SFT}
    \label{fig-example-ft}
    \end{figure}

@@

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
definition productT :: |$"('q,'a,'i,'b)"$| NFT |$\Rightarrow$| |$('q,'a)$| NFA |$\Rightarrow$| 
           |$(('a,'b)$| Tlabel |$\Rightarrow 'a \text{ set}\Rightarrow 'b\text{ set option})\Rightarrow$|
           |$('q\times 'q, ~'b)$| eNFA where
  "productT |$\mathcal{T}$| |$\mathcal{A}$| F = |$\llparenthesis$|
    |$\mathcal{Q}e = \mathcal{Q}_t~\mathcal{T} \times \mathcal{Q} ~\mathcal{A}$|,
    |$\Delta_e = \{((p,p'),~\texttt{the }(((\mathcal{M}_t~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A}~\land$|
          |$(p, (\texttt{None}, f), q)\in \Delta_t~\mathcal{T} \land \exists S.~(\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{Some } S\} ~\cup$|
        |$\{((p,p'),~\texttt{the }(F~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 \neq \emptyset~\land$|
            |$ \exists S.~F~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2) = \text{Some } S\}$|,
    |$\Delta_e' = \{((p,p'),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
          |$(p, (None, f), q)\in \Delta_t~\mathcal{T} \land (\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{None}\} ~\cup$|
        |$\{((p,p'),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 \neq \emptyset~\land$|
            |$ \exists x\in(\sigma_1\cap\sigma_2).~((\mathcal{M}_t~\mathcal{T})~f)~(\text{Some } x) = \text{None} \}$|,
    |$\mathcal{I}_e = \mathcal{I}_t~\mathcal{T} \times \mathcal{I}~\mathcal{A}$|
    |$\mathcal{F}_e = \mathcal{F}_t~\mathcal{T} \times \mathcal{F}~\mathcal{A}$| |$\rrparenthesis$|"
	\end{lstlisting}
\caption{The formalization of product operation}
\label{fig-def-FTProd}
\end{figure}

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
fun inputE :: |$('a\;\text{option}\;\times\;'b\;\text{option})\;\text{list}\;\Rightarrow\;'a\;\text{list}$| where
  "inputE [] = []" |$\mid$|
  "inputE ((Some a, |$\_$|) |$\#$| l) = a |$\#$| (inputE l)" |$\mid$|
  "inputE ((None, |$\_$|) |$\#$| l) = (inputE l)"

fun outputE :: "|$('a \text{ option} \times 'b \text{ option}) \text{ list} \Rightarrow 'b \text{ list}$|" where
  "outputE [] = []" |$\mid$|
  "outputE ((_, Some a) # l) = a # (outputE l)" |$\mid$|
  "outputE ((_, None) # l) = (outputE l)"

definition outputL :: "|$('q, 'a, 'i, 'b)\text{  NFT} \Rightarrow 
                               ('q, 'a) \text{ NFA} \Rightarrow 'b\text{ list set}$|" where
  "outputL |$\mathcal{T}$| |$\mathcal{A}$| = {outputE |$\pi~\mid~\pi~q~q'$|. |$q\in \mathcal{I}_t~\mathcal{T} \land q' \in \mathcal{F}_t~\mathcal{T} \land$|
                |$ \text{LTTS\_reachable }\mathcal{T}~q~\pi~q'\land \text{inputE }\pi\in \mathcal{L}~\mathcal{A}$|}" 
	\end{lstlisting}
\caption{The formalization of traces in SFTs}
\label{fig-def-output}
\end{figure}

\begin{figure}[hbt!]
	\begin{lstlisting}
lemma productT_correct:
  fixes |$\mathcal{T}$| |$\mathcal{A}$| F
  assumes F_ok1: "|$\forall~f~s.~(\forall e \in s.~f~(\text{Some } e) = \text{None}) \longleftrightarrow\text{ F}~f~s = \text{None}"$|
      and F_ok2: "|$\forall~f~s.~ \texttt{F}~f~s\neq \texttt{None} \longrightarrow$| F |$f~s$| = |$\text{Some }(\cup~\{\texttt{S} \mid e~\texttt{S}.~e \in s \land f~(\text{Some }e) = \text{ Some S}\})$|"
      and wfTA: "NFT_wf |$\mathcal{T}~\land \text{ NFA } \mathcal{A}$|"
    |$\textcolor{blue}{\text{shows}}$| "|$\mathcal{L}_e$| (productT |$\mathcal{T}~\mathcal{A}$| F) = outputL |$\mathcal{T}~\mathcal{A}$|"

@@

To facilitate formal reasoning and optimize performance, we introduce a canonical form for interval. An interval $[(i_1, j_1), \ldots, (i_n, j_n)]$ is in canonical form if it satisfies two key properties:
\begin{enumerate}
  \item Each $(i_k, j_k)$ is well-formed: $i_k \leq j_k$ for all $k \in \{1,\ldots,n\}$
  \item Intervals are ordered and non-overlapping: $j_k < i_{k+1}$ for all $k \in \{1,\ldots,n-1\}$
\end{enumerate}

@@

In this subsection, we present the algorithmic implementation of the product operation between an SFT and an SFA\footnote{For clarity of presentation, we show a simplified version of the Isabelle/HOL implementation while preserving the essential algorithmic structure.}. 
%
Figure \ref{fig-compute-nft-product} illustrates the core algorithm \texttt{productT\_impl}, which is implemented using the \emph{Refine\_Monadic} framework. Note that in the implementation, there are some refined operations: \texttt{nfa\_states}, \texttt{nfa\_trans}, \texttt{nfa\_initial}, \texttt{nfa\_accepting}, and \texttt {nft\_tranfun}. These are corresponding to the states, transitions, initial states, accepting states, and output function mapping of the SFT.

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
definition trans_comp_imp where
  "trans_comp_imp M F fe T1 T2 Q =
    FOREACH {t. t |$\in$| T1}
      (|$\lambda$| (q, (|$\alpha$|, f), q|$'$|) (D1, D2). 
        (if (|$\alpha$| = None) then 
          (subtrans_comp_|$\varepsilon$| M q f q|$'$| F fe T2 D1 D2)
        else 
          (subtrans_comp M q (the |$\alpha$|) f q|$'$| F fe T2 D1 D2)
      )) (|$\emptyset$|, |$\emptyset$|)"
\end{lstlisting}
\caption{The computation of \texttt{trans\_comp\_imp}}
\label{fig-def-prods-imp}
\end{figure}

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
definition subtrans_comp where
  "subtrans_comp M q |$\alpha$| f q|$'$| F fe T D1 D2 =
    FOREACH
      {t. t |$\in$| T} (|$\lambda$| (q1, |$\alpha'$|, q1|$'$|) (D1, D2).
        (if (nemptyIs (intersectIs |$\alpha$| |$\alpha'$|)) then do {
          D1 |$\leftarrow$| (if (F (M f) (intersectIs |$\alpha$| |$\alpha'$|)) |$\neq$| None) 
            then  
            let |$\alpha_i$| = the (F (M f) (intersectIs |$\alpha$| |$\alpha'$|)) in
            RETURN {((q,q1), |$\alpha_i$|, (q|$'$|,q1|$'$|))} |$\cup$| D1
            else RETURN D1);
          D2 |$\leftarrow$| (if fe (M f) (intersectIs |$\alpha$| |$\alpha'$|) then 
            RETURN {((q,q1), (q|$'$|,q1|$'$|))} |$\cup$| D2 else RETURN D2);
          RETURN (D1, D2)
        }
        else (RETURN (D1, D2)))) (D1, D2)"
    \end{lstlisting}
    \caption{The computation of \texttt{subtrans\_comp}}
    \label{fig-def-subtrans_comp}
    \end{figure}

@@

The string replacement operation, denoted as $\texttt{replace}(\texttt{str}, \texttt{pattern}, \texttt{replacement})$, is a fundamental string transformation that takes three parameters:
\begin{itemize}
  \item $\texttt{str}$: The input string to be transformed.
  \item $\texttt{pattern}$: A regular expression defining the matching criteria.  
  \item $\texttt{replacement}$: The string to be substituted for the matched substring.
\end{itemize}

@@

The first case can be modeled by SFTs. We illustrate this case by an example. Given a replacement operation $\texttt{replace}(s, $\texttt{/[0-9]+/}$, $"$\texttt{NUM}$"$)$, which means replacing a occurrence of the substring that matches the regular expression $\texttt{/[0-9]+/}$ with the string "\texttt{NUM}".
We model this replacement operation by an SFT with the following 3 steps:

>>>>

\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\title{Certified Symbolic Finite Transducers: Formalization and Applications to String Manipulation}

\begin{abstract}
Finite Transducers (FTs) extend the capabilities of Finite Automata (FAs) by enabling the transformation of input strings into output strings, thereby providing a more expressive framework for operations that encompass both recognition and transformation. FTs have wide applications, including program analysis, string constraint solving, and security-critical sanitizer analysis.
However, there is currently no scalable formalization of FTs. This is primarily because transition labels in FTs are not formalized symbolically, resulting in transition explosion and significant performance bottlenecks in practical applications.

In this paper, we present a formalization of FTs in the symbolic setting using Isabelle/HOL, where transition labels can be instantiated with various Boolean algebras, such as intervals and arithmetic predicates. This symbolic approach enables not only efficient FT operations—such as computing the output language of an FT given a regular language—but also supports infinite alphabets.
%
To evaluate the effectiveness of our formalization, we apply the formalized SFTs to two applications: (1) \texttt{HTMLdecode}, a sanitizer used for preventing XSS attacks, and (2) a string solver that models replacement operations with various semantics, including left-most and replace-all matching. Experimental results demonstrate that our approach achieves computational efficiency in constraint-solving scenarios.
\end{abstract}

@@

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite transducers in Isabelle/HOL proof assistant.
\item We leverage the refinement framework to create an extensible SFT formalization that is capable of accommodating various boolean algebras, ensuring adaptability to future transition label developments.
\item We develop an interval algebra with efficient set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization through its application to string constraint solving, specifically in modeling replacement operations with verified correctness guarantees.
\end{enumerate}

@@

For a word $w = a_0,\ldots, a_{n-1}$, it is accepted by $\sigma$ if and only if $a_i \in \llbracket\phi_i \rrbracket$ for $0 \leq i \leq n - 1$. When $w$ is accepted, run $\sigma$ generates an output sequence $w'= f_0~a_0, \ldots, f_{n-1}~a_{n-1}$. We define:
\begin{itemize}
\item  $(a_0,(f_0~a_0)), \ldots, (a_{n-1},(f_{n-1}~a_{n-1}))$ a \emph{trace},
\item $a_0,\ldots,a_{n-1}$ the \emph{input} of the trace and $(f_0~a_0), \ldots, $ $(f_{n-1}~a_{n-1})$ the \emph{output} of the trace.
\end{itemize}

If $a_0,\ldots,a_{n-1}$ is accepted by an accepting run in $\mathcal{T}$, we say that the trace is an accepting trace of $\mathcal{T}$.
For a trace $\pi$, we denote its input as $\mathit{in}(\pi)$ and its output as $\mathit{out}(\pi)$.
%
Given an SFT $\mathcal{T}$ and a word $w$, we define the \emph{product} operation of $\mathcal{T}$ and $w$ (denoted as $\mathcal{T}\times\{w\}$) as the set of outputs generated by $\mathcal{T}$ with input $w$. More precisely,
\[
\begin{split}
\mathcal{T}\times\{w\} = \{w'\mid \exists \pi.~\pi \text{ is an accepting trace of } \mathcal{T} \land \\mathit{in}(\pi) = w \land \mathit{out}(\pi) = w'\}.
\end{split}
\]

To make the operation \emph{product} more general, we extend the operation to an SFT and a set of input words represented by a regular language, which can be denoted by an SFA $\mathcal{A}$. More precisely,
\[
\begin{split}
\mathcal{T}\times \mathcal{A} = \{w'\mid \exists w.~w\in \mathcal{L}(\mathcal{A})\land w' \in \mathcal{T}\times\{w\}\}, \\\text{ where }\mathcal{L}(\mathcal{A})\text{ denotes the language of }\mathcal{A}.
\end{split}
\]

\begin{figure}[hbt!]
  \centering
  \begin{tikzpicture}[shorten >=1pt, node distance=2.5cm and 2cm, on grid, auto]
    % Make q0 and q1 horizontally wide apart
    \node[state, initial] (q_0)   {$q_0$}; 
    \node[state] (q_1) [right=4cm of q_0] {$q_1$}; 
    \node[state, accepting] (q_2) [below=2.2cm of q_1] {$q_2$};

    \path[->] 
    (q_0) edge node {[$'\texttt{a}'-'\texttt{z}'$],$f_1$} (q_1)
    (q_1) edge node[right] {[$'\texttt{A}'-'\texttt{Z}'$],$f_2$} (q_2)
    (q_1) edge[loop above] node {[$'\texttt{a}'-'\texttt{z}'$],$f_1$} (q_1)
    (q_2) edge[loop below] node {[$'\texttt{A}'-'\texttt{Z}'$],$f_2$} (q_2);
  \end{tikzpicture}
    \caption{An example of SFT}
    \label{fig-example-ft}
\end{figure}

@@

Firstly, we present the basic idea of our formalization of SFTs in Isabelle/HOL, which consists of three layers.

\begin{figure}[hbt!]
\centering
\begin{tikzpicture}[
  node distance=1cm,
  every node/.style={font=\small},
  layer/.style={rectangle, draw, rounded corners, fill=blue!10, minimum width=5.5cm, minimum height=1.2cm, align=center},
  arrow/.style={->, thick}
]

% Abstract Layer
\node[layer] (abstract) {%
  \textbf{Abstract Layer}\\
  FT with transition labels as \textbf{sets}
};

% Locale Layer
\node[layer, below=of abstract] (locale) {%
  \textbf{Implementation Layer}\\
  FT with transition labels as a \\boolean algebra locale
};

% Implementation Layer
\node[layer, below=of locale] (impl) {%
  \textbf{Boolean Algebra Layer}\\
  Concrete Boolean algebra implementations\\
  (e.g., intervals, bitvectors, predicates)
};

% Draw arrows with refinement labels
\draw[arrow] (abstract) -- node[right] {refinement} (locale);
\draw[arrow] (locale) -- node[right] {refinement} (impl);

\end{tikzpicture}
\caption{Three-layered formalization of SFTs in Isabelle/HOL. 
\textbf{Abstract Layer:} Transition labels are modeled as sets, representing the most general case.
\textbf{Locale Layer:} Transition labels are abstracted as a type variable \texttt{'a} within a locale, supporting any Boolean algebra.
\textbf{Implementation Layer:} Concrete Boolean algebra instances, such as intervals, bitvectors, or predicates, are provided.}
\label{fig:layered-formalization}
\end{figure}

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
definition productT :: |$"('q,'a,'i,'b)"$| NFT |$\Rightarrow$| 
   |$('q,'a)$| NFA |$\Rightarrow$||$(('a,'b)$| Tlabel |$\Rightarrow 'a \text{ set}$| 
   |$\Rightarrow 'b\text{ set option})\Rightarrow$| |$('q\times 'q, ~'b)$| eNFA where
"productT |$\mathcal{T}$| |$\mathcal{A}$| F = |$\llparenthesis$|
  |$\mathcal{Q}e = \mathcal{Q}_t~\mathcal{T} \times \mathcal{Q} ~\mathcal{A}$|,
  |$\Delta_e = \{((p,p'),~\texttt{the }(((\mathcal{M}_t~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid $|
       |$p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A}~\land$| |$(p, (\texttt{None}, f), q)\in \Delta_t~\mathcal{T}$|
        |$ \land~\exists S.~(\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{Some } S\} ~\cup$|
      |$\{((p,p'),~\texttt{the }(F~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),(q, q'))\mid $|
       |$p, p', q, \sigma_1, \sigma_2, q', f.~(p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land$|
       |$ (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 \neq \emptyset~\land$|
          |$ \exists S.~F~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2) = \text{Some } S\}$|,
  |$\Delta_e' = \{((p,p'),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
       |$(p, (None, f), q)\in \Delta_t~\mathcal{T}~\land $|
       |$(\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{None}\} ~\cup$|
      |$\{((p,p'),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
        |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}$|
        |$\land~\sigma_1\cap \sigma_2 \neq \emptyset~\land\exists x\in(\sigma_1\cap\sigma_2).$|
        |$~((\mathcal{M}_t~\mathcal{T})~f)~(\text{Some } x) = \text{None} \}$|,
  |$\mathcal{I}_e = \mathcal{I}_t~\mathcal{T} \times \mathcal{I}~\mathcal{A}$|
  |$\mathcal{F}_e = \mathcal{F}_t~\mathcal{T} \times \mathcal{F}~\mathcal{A}$| |$\rrparenthesis$|"
	\end{lstlisting}
\caption{The formalization of product operation}
\label{fig-def-FTProd}
\end{figure}

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
fun inputE :: 
   "|$('a\;\text{option}\;\times\;'b\;\text{option})\;\text{list}\;\Rightarrow\;'a\;\text{list}$|"
where
  "inputE [] = []" |$\mid$|
  "inputE ((Some a, |$\_$|) |$\#$| l) = 
                      a |$\#$| (inputE l)" |$\mid$|
  "inputE ((None, |$\_$|) |$\#$| l) = (inputE l)"

fun outputE :: 
    "|$('a \text{ option} \times 'b \text{ option}) \text{ list} \Rightarrow 'b \text{ list}$|" 
where
  "outputE [] = []" |$\mid$|
  "outputE ((_,Some a) # l) = 
                    a # (outputE l)" |$\mid$|
  "outputE ((_,None) # l) = (outputE l)"

definition outputL :: 
   "|$('q, 'a, 'i, 'b)\text{  NFT} \Rightarrow 
                               ('q, 'a) \text{ NFA} \Rightarrow 'b\text{ list set}$|" where
  "outputL |$\mathcal{T}$| |$\mathcal{A}$| = {outputE |$\pi~\mid~\pi~q~q'$|. 
    |$q\in \mathcal{I}_t~\mathcal{T} \land q' \in \mathcal{F}_t~\mathcal{T}~\land$| 
    |$ \text{LTTS\_reachable }\mathcal{T}~q~\pi~q'\land$||$ \text{inputE }\pi\in \mathcal{L}~\mathcal{A}$|}"
	\end{lstlisting}
\caption{The formalization of traces in SFTs}
\label{fig-def-output}
\end{figure}

\begin{figure}[hbt!]
	\begin{lstlisting}
lemma productT_correct:
  fixes |$\mathcal{T}$| |$\mathcal{A}$| F
  assumes 
   F_ok1: "|$\forall~f~s.~(\forall e \in s.~f~(\text{Some } e) = \text{None})\longleftrightarrow$|
       |$\text{ F}~f~s = \text{None}"$|
   and F_ok2: "|$\forall~f~s.~ \texttt{F}~f~s\neq \texttt{None} \longrightarrow$| F |$f~s$| = 
        |$\text{Some }(\cup~\{\texttt{S} \mid e~\texttt{S}.~e \in s~\land~$|
        |$f~(\text{Some }e) = \text{ Some S}\})$|"
      and wfTA: "NFT_wf |$\mathcal{T}~\land \text{ NFA } \mathcal{A}$|"
  |$\textcolor{blue}{\text{shows}}$| "|$\mathcal{L}_e$| (productT |$\mathcal{T}~\mathcal{A}$| F) = outputL 
          |$\mathcal{T}~\mathcal{A}$|"

@@

To facilitate formal reasoning and optimize performance, we introduce a canonical form for interval. An interval $[(i_1, j_1), $ $\ldots, (i_n, j_n)]$ is in canonical form if it satisfies two key properties:
\begin{enumerate}
  \item Each $(i_k, j_k)$ is well-formed: $i_k \leq j_k$ for all $k \in \{1,\ldots,n\}$
  \item Intervals are ordered and non-overlapping: $j_k < i_{k+1}$ for all $k \in \{1,\ldots,n-1\}$
\end{enumerate}

@@

In this subsection, we present the algorithmic implementation of the product operation between an SFT and an SFA\footnote{For clarity of presentation, we show a simplified version of the Isabelle/HOL implementation while preserving the essential algorithmic structure.}. 
%
Figure \ref{fig-compute-nft-product} illustrates the core algorithm \texttt{productT\_impl}, which is implemented using the \emph{Refine\_Monadic} framework. Note that in the implementation, there are some refined operations: \texttt {nft\_tranfun}, \texttt{nfa\_states}, \texttt{nfa\_trans}, \texttt{nfa\_initial}, and \texttt{nfa\_accepting}. These are corresponding to the states, transitions, initial states, accepting states, and output function mapping of the SFT.

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
definition trans_comp_imp where
"trans_comp_imp M F fe T1 T2 Q =
  FOREACH {t. t |$\in$| T1}
   (|$\lambda$|(q,(|$\alpha$|,f),q|$'$|) (D1,D2). 
    (if (|$\alpha$|=None) then 
     (subtrans_comp_|$\varepsilon$| M q f q|$'$| F fe 
                              T2 D1 D2)
     else
     (subtrans_comp M q (the |$\alpha$|) f 
            q|$'$| F fe T2 D1 D2))) (|$\emptyset$|, |$\emptyset$|)"
\end{lstlisting}
\caption{The computation of \texttt{trans\_comp\_imp}}
\label{fig-def-prods-imp}
\end{figure}

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
definition subtrans_comp where
"subtrans_comp M q |$\alpha$| f q|$'$| F fe T D1 D2 =
 FOREACH {t.t|$\in$|T} (|$\lambda$| (q1,|$\alpha'$|,q1|$'$|) (D1,D2).
  (if (nemptyIs (intersectIs |$\alpha$| |$\alpha'$|)) then
    do {
      D1 |$\leftarrow$| 
       (if (F (M f) (intersectIs |$\alpha$| |$\alpha'$|)) 
          |$\neq$| None) then
         let |$\alpha_i$| = the (F (M f) 
              (intersectIs |$\alpha$| |$\alpha'$|)) in
         RETURN {((q,q1), |$\alpha_i$|, (q|$'$|,q1|$'$|))} 
                |$\cup$| D1
        else RETURN D1);
      D2 |$\leftarrow$| 
       (if fe (M f) (intersectIs |$\alpha$| |$\alpha'$|) 
        then 
         RETURN {((q,q1), (q|$'$|,q1|$'$|))} |$\cup$| D2 
        else RETURN D2);
      RETURN (D1, D2) }
   else (RETURN (D1, D2)))) (D1, D2)"
    \end{lstlisting}
    \caption{The computation of \texttt{subtrans\_comp}}
    \label{fig-def-subtrans_comp}
    \end{figure}

@@

The string replacement operation, denoted as 
$\texttt{replace}(\texttt{str}, \newline\texttt{pattern}, \texttt{replacement})$, is a fundamental string transformation that takes three parameters:
\begin{itemize}
  \item $\texttt{str}$: The input string to be transformed.
  \item $\texttt{pattern}$: A regular expression defining the matching criteria.  
  \item $\texttt{replacement}$: The string to be substituted for the matched substring.
\end{itemize}

@@

The first case can be modeled by SFTs. We illustrate this case by an example. Given a replacement operation $\texttt{replace}(s, \texttt{/[0-9]+/}, "\texttt{NUM}")$, which means replacing a occurrence of the substring that matches the regular expression $\texttt{/[0-9]+/}$ with the string "\texttt{NUM}".
We model this replacement operation by an SFT with the following 3 steps:
